---
title: "Estimating Flood Inundation Probability Map"
author: "Lu Yii Wong & Yujin Song"
date: "2024-03-22"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# 1: Setup
```{r setup, warning = FALSE, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(pscl)
library(plotROC)
library(pROC)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tigris)
library(viridis)
library(ggplot2)
library(dplyr)
```

```{r themes}
mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.75),
  axis.ticks=element_blank())

```

## 1.1: Load Data
```{r load data}
# dependent variable
inundation <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_depe.csv")

# independent variables
elevation <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_elev.csv")

flow_acc <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_fac.csv")

land_use <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_lulc.csv")

slope <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_slope.csv")

water_body <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_water.csv")

```

# 2: Exploratory Analysis
## 2.1: Fishnet and Joining to show Past Flood Inundation
```{r Fishnet}

calgary_fishnet<-st_read("https://raw.githubusercontent.com/luyiiw/LUEM/main/Calgary_JSON/Calgary_Fishnet.json")
calgary_boundary<-st_read("https://raw.githubusercontent.com/luyiiw/LUEM/main/Calgary_JSON/Calgary_Boundary.json")

calgary_Floods <- calgary_fishnet %>% 
  left_join(inundation, by = "FID")

calgary_Floods$RANGE<-as.factor(calgary_Floods$RANGE)

ggplot() + 
  geom_sf(data=calgary_fishnet,color="grey",size=0.2) +
  geom_sf(data=calgary_Floods ,aes(fill=RANGE),size=0.5) +
   scale_fill_manual(values = c("1"= "#1f78b4", "0" = "#e7f5c6"),
                     labels = c("1" = "Inundation", "0" = "No Inundation"),
                  name = "Inundation Status") +
  labs(title="Past Flood Inundation in Calgary, Canada",
       caption = "Figure 1.1")+
  mapTheme+
    guides(fill = guide_legend(override.aes = list(shape = 22, size = 6)))

```

## 2.2: Data Exploration
```{r}
# Creating tables to combine exploratory data
flood_dep <- inundation %>%
  select(FID, RANGE) %>%
  rename(flood = RANGE) %>%
  mutate(flood = factor(flood))

# 1 = flood, 0 = no flood

#1 Elevation
elev_comb <- elevation %>%
  select(FID, MEAN) %>%
  rename(mean_elev= MEAN)

#2 Slope
slope_comb <- slope %>%
  select(FID, MAX) %>%
  rename(max_slope= MAX)

#3 Water body
water_comb <- water_body %>%
  select(FID, MIN) %>%
  rename(min_dist_water= MIN)

#4 Flow accumulation
fac_comb <-  flow_acc %>%
  select(FID, MEAN) %>%
  rename(mean_fac= MEAN)
#changed from max to mean

#5 Land Use/Permeability
lu_comb <- land_use %>%
  select(FID, MEAN) %>%
  rename(land_use = MEAN) %>%
  mutate(permeability = case_when(
    land_use < 0.25 ~ '0',
    land_use >= 0.25 ~ '1')) %>%
  select(-land_use)

# Removing unnecessary data sets
rm(inundation, elevation, flow_acc, land_use, slope, water_body)

# Combining the tables 
flood_comb <- flood_dep %>%
  inner_join(elev_comb, by = "FID") %>%
  inner_join(slope_comb, by = "FID") %>%
  inner_join(water_comb, by = "FID") %>%
  inner_join(fac_comb, by = "FID")

floodPlotVariables <- 
  flood_comb %>%
  as.data.frame() %>%
  select(flood, mean_elev, max_slope, min_dist_water, mean_fac) %>%
  gather(variable, value, -flood)

```

```{r plot_variables}
# need to check the labels for flood and not flood 
ggplot(floodPlotVariables %>%
         group_by(flood, variable) %>%
         summarize(mean = mean(value))) + 
     geom_bar(aes(as.factor(flood), 
                  mean, 
                  fill=as.factor(flood)),
              stat="identity") + 
     facet_wrap(~variable, scales = "free_y") +
     scale_fill_manual(values = c("#e7f5c6", "#1f78b4"),
                      labels = c("No Flood","Flood"),
                      name = "") +
  labs(title="Continuous Variables Impacting Flood Risk",
       caption = "Figure 2.1")+
    labs(x="Flood", y="Value") +
  plotTheme

```


# Model Building 
## Partition Training and Test Sets
```{r}
# Combine all data sets for model building
model_data <- flood_dep %>%
  inner_join(elev_comb, by = "FID") %>%
  inner_join(slope_comb, by = "FID") %>%
  inner_join(water_comb, by = "FID") %>%
  inner_join(fac_comb, by = "FID") %>%
  inner_join(lu_comb, by = "FID") %>%
  mutate(permeability = as.factor(permeability))

#land cover is dummy var and categorical var


# Creating training and test sets
set.seed(3456)
trainIndex <- createDataPartition(model_data$permeability, p = .70,
                                  list = FALSE,
                                  times = 1)

floodTrain <- model_data[ trainIndex,]
floodTest  <- model_data[-trainIndex,]

```


## Making the Binomial Model
```{r}
floodModel <- glm(flood ~ ., 
                    family="binomial"(link="logit"), data = floodTrain %>%
                                                            as.data.frame() %>%
                                                            select(-FID))
summary(floodModel)
```

## Model Validation
```{r}
classProbs <- predict(floodModel, floodTest, type="response")

hist(classProbs, col = "#1f78b4", main = "Histogram of Flood Probabilities") 


testProbs <- data.frame(obs = as.factor(floodTest$flood),
                        pred = classProbs)

ggplot(testProbs, aes(x = pred, fill=as.factor(obs))) + 
  geom_density() +
  facet_grid(obs ~ .) + 
  xlab("Probability") +
  ylab("Frequency")+
  geom_vline(xintercept = .5) +
  labs(title="Model Validation",
       caption = "Figure 3.2")+
  scale_fill_manual(values = c("#e7f5c6", "#1f78b4"),
                      labels = c("Not Flood","Flood"),
                      name = "")+
  plotTheme
```

## Confusion Matrix
```{r}
testProbs$predClass  = ifelse(testProbs$pred > .5 ,1,0)

caret::confusionMatrix(reference = as.factor(testProbs$obs), 
                       data = as.factor(testProbs$predClass), 
                       positive = "1")
```

Predicted = 0, Observed = 0 —> True Negative (No flood correctly predicted 130)

Predicted = 1, Observed = 1 —> True Positive (Floods correctly predicted 40)

Predicted = 1, Observed = 0 —> False Positive (Incorrectly predicted flood when there was no flood 23)

Predicted = 0, Observed = 1 —> False Negative (Incorrectly predicted no flood when there was a flood 10)

Sensitivity: 0.63
Specificity: 0.92


## ROC Curve

```{r roc_curve, message = FALSE, warning = FALSE}
# change obs to numeric
testProbs$obs <- as.numeric(as.character(testProbs$obs))

# plot results
ggplot(testProbs, aes(d = obs, m = pred)) + 
  geom_roc(n.cuts = 50, labels = FALSE) + 
  style_roc(theme = theme_grey) +
  geom_abline(slope = 1, intercept = 0, size = 1.5, color = 'grey') +
  labs(title="ROC Curve",
       caption = "Figure 3.3") +
  plotTheme

#area under the curve
auc(testProbs$obs, testProbs$pred)
```
Area under the curve: 0.85

## Cross-Validation
```{r}
ctrl <- trainControl(method = "cv", 
                     number = 100, 
                     p = 0.7, 
                     savePredictions = TRUE)

cvFit <- train(as.factor(flood) ~ .,  data = model_data %>% 
                                                as.data.frame(), 
               method="glm", family="binomial",
               trControl = ctrl)

cvFit

ggplot(as.data.frame(cvFit$resample), aes(Accuracy)) + 
  geom_histogram(fill = "#1f78b4", color = "black") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x="Accuracy",
       y="Count") +
  labs(title="Cross-Validation",
       caption = "Figure 3.4")+
  plotTheme

```
Accuracy: 0.86
Kappa: 0.63

## Map Predictions
```{r}
###Refine here by yujin
allPredictions <- 
  predict(cvFit, model_data, type="prob")[,2]
  
finalflood <- 
  cbind(model_data,allPredictions) %>%
  mutate(allPredictions = round(allPredictions * 100)) 

joined_fishnet_cal <- calgary_fishnet %>% 
  left_join(finalflood, by = "FID")

ggplot() + 
    geom_sf(data=joined_fishnet_cal, aes(fill=factor(ntile(allPredictions,5))), 
            colour=NA) +  # Raster layer
  geom_sf(data = calgary_fishnet, color = "grey", fill = NA) + 
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels=as.character(quantile(joined_fishnet_cal$allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  mapTheme +
  labs(title="Predicted Probability of Flood Inundation in Calgary, Canada")



ggplot() + 
    geom_sf(data=joined_fishnet_cal, aes(fill=factor(ntile(allPredictions,5))), 
            colour=NA) +
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels=as.character(quantile(joined_fishnet_cal$allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  geom_sf(data=joined_fishnet_cal  %>% 
               filter(flood == 1), 
               fill="#de2d26",colour=NA) +
  geom_sf(data = calgary_fishnet, color = "grey", fill = NA) + 
  mapTheme +
  labs(title="Predicted Probabilities and Past Flood Inundation",
       subtitle="Calgary; Past Flood Inundation in red")


##Confusion Matrix Map
joined_fishnet_cal %>%
  mutate(confResult=case_when(allPredictions < 50 & flood==0 ~ "True Negative",
                              allPredictions >= 50 & flood==1 ~ "True Positive",
                              allPredictions < 50 & flood==1 ~ "False Negative",
                              allPredictions >= 50 & flood==0 ~ "False Positive")) %>%
  ggplot()+
  geom_sf(aes(fill = confResult), color = "transparent")+
   geom_sf(data = calgary_fishnet, color = "grey", fill = NA) +
  scale_fill_manual(values = c("Red","Orange","Light Blue","Light Green"),
                    name="Outcomes")+
  labs(title="Confusion Matrix Map") +
  mapTheme
```


# APPLY MODEL TO DENVER
## Load Data
```{r load data for Denver}
# independent variables
elevation_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_elevation_den.csv")

flow_acc_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_fac_den.csv")

land_use_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_permea_den.csv")

slope_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_slope_den.csv")

water_body_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_waterdist_den.csv")

```

## Denver data cleaning
```{r}
#1 Elevation
elev_comb_den <- elevation_den %>%
  select(FID, MEAN) %>%
  rename(mean_elev= MEAN)

#2 Slope
slope_comb_den <- slope_den %>%
  select(FID, MAX) %>%
  rename(max_slope= MAX)

#3 Water body
water_comb_den <- water_body_den %>%
  select(FID, MIN) %>%
  rename(min_dist_water= MIN)

#4 Flow accumulation
fac_comb_den <-  flow_acc_den %>%
  select(FID, MEAN) %>%
  rename(mean_fac= MEAN)
#changed from max to mean

#5 Land Use/Permeability
lu_comb_den <- land_use_den %>%
  select(FID, MEAN) %>%
  rename(land_use = MEAN) %>%
  mutate(permeability = case_when(
    land_use < 0.25 ~ '0',
    land_use >= 0.25 ~ '1')) %>%
  select(-land_use)

```

## Model Building Denver
```{r}
# Combine all data sets for DENVER model building
model_data_den <- elev_comb_den %>%
  inner_join(slope_comb_den, by = "FID") %>%
  inner_join(water_comb_den, by = "FID") %>%
  inner_join(fac_comb_den, by = "FID") %>%
  inner_join(lu_comb_den, by = "FID") %>%
  mutate(permeability = as.factor(permeability))

# Removing unnecessary data sets 
rm(elev_comb_den, slope_comb_den, water_comb_den, fac_comb_den, lu_comb_den)
```

## Predicting Flood in Denver
```{r}
den_classProbs <- predict(floodModel, model_data_den, type = "response")

den_floodModel <- cbind(model_data_den, den_classProbs) %>%
  mutate(allPredictions2 = round(den_classProbs * 100)) 

joined_fishnet_den <- denver_fishnet %>% 
  left_join(den_floodModel , by = "FID")

ggplot() + 
    geom_sf(data=joined_fishnet_den, aes(fill=factor(ntile(allPredictions2,5))), 
            colour=NA) +  # Raster layer
  geom_sf(data = denver_fishnet, color = "grey", fill = NA) + 
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels=as.character(quantile(joined_fishnet_den$allPredictions2,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  mapTheme +
  labs(title="Predicted Probability of Flood Inundation in Denver, Colorado")

breaks_values <- quantile(joined_fishnet_den$allPredictions2, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
labels_values <- format(breaks_values, digits = 4, nsmall = 2)

ggplot() + 
    geom_sf(data = joined_fishnet_den, aes(fill = factor(ntile(allPredictions2, 5))), colour = NA) +
    geom_sf(data = denver_fishnet, color = "grey", fill = NA) +
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels = labels_values,
                      name = "Predicted Probabilities (%)") +
    labs(title = "Predicted Probability of Flood Inundation in Denver, Colorado") +
    mapTheme


breaks <- quantile(joined_fishnet_den$allPredictions2, probs = seq(0, 1, 0.25), na.rm = TRUE)
labels <- formatC(breaks, format = "f", digits = 2)  # Use formatC for more control over formatting

ggplot() + 
    geom_sf(data = joined_fishnet_den, aes(fill = factor(ntile(allPredictions2, 5))), colour = NA) +  # Raster layer
    geom_sf(data = denver_fishnet, color = "grey", fill = NA) +  # Denver fishnet overlay
    scale_fill_manual(
      values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),  # Slightly revised for clearer color progression
      labels = labels,  # Use formatted labels for clarity
      name = "Predicted\nProbabilities(%)\n(Quintile\nBreaks)"
    ) +
    labs(title = "Predicted Probability of Flood Inundation in Denver, Colorado") +
  mapTheme


quantile_breaks <- quantile(joined_fishnet_den$allPredictions2, probs = seq(0, 1, by = 0.2), na.rm = TRUE)
# Create custom labels based on these breaks
custom_labels <- sprintf("%.2f - %.2f", head(quantile_breaks, -1), tail(quantile_breaks, -1))

ggplot() + 
    geom_sf(data = joined_fishnet_den, aes(fill = factor(ntile(allPredictions2, 5))), colour = NA) +
    geom_sf(data = denver_fishnet, color = "grey", fill = NA) +
    scale_fill_manual(
      values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),  # Your color choices
      labels = custom_labels,  # Use the custom labels based on quantile breaks
      name = "Predicted\nProbabilities(%)\n(Quintile\nBreaks)"
    ) +
    labs(title = "Predicted Probability of Flood Inundation in Denver, Colorado") +
  mapTheme
  # Or any other theme or customization you prefer




```

