---
title: "Estimating Flood Inundation Probability Map"
author: "Lu Yii Wong & Yujin Song"
date: "2024-03-22"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# 1: Introduction
## 1.1: Context of Flooding in Calgary
The city of Calgary is a located in Alberta, Canada. According to the national census, Calgary had a population of 1,306,784 in 2021 [Statistics Canada](https://www12.statcan.gc.ca/census-recensement/2021/search-recherche/productresults-resultatsproduits-eng.cfm?LANG=E&GEOCODE=2021A00054806016). Calgary has a dry climate, however, because of its proximity to mountains, the city is vulnerable to “unpredictable swings in the weather from heavy rains leaving to floods to many weeks of dry temperatures and little rain leading to droughts.” [Calgary.ca](https://www.calgary.ca/water/flooding.html) 

In 2013, the city experienced a devastating flood. The flood was a result of “melting snowpack in the Rocky mountains combined with steep, rocky terrain caused rapid and intense flooding in southern-Alberta watersheds.” [Calgary.ca](https://www.calgary.ca/water/flooding/history-calgary.html) This flood resulted in one fatality, injuries, and massive infrastructure damage. The estimated damages from this flooding event is over $6 billion across the southern Alberta. This natural disaster resulted in tremendous infrastructure loss, financial burden in recovery, and trauma by Calgary residents impacted. 

## 1.2: Planning Motivation
Our team believes that studying flood inundation risk is important to prepare our cities from future impacts of flooding. Modeling flood inundation will inform planners and policy makers on high-flood risk areas. With this knowledge we can plan for the future, by recommending plans and policies to curb future development in high risk regions and protect moderate risk areas. 

For this exercise, our team used Calgary as our case study to examine physical features that are correlated to flood risk. Based on our data exploration, we built a model to predict future flooding in Calgary, tested model performance, and applied it to a comparable city. Additionally, we chose Denver a city in Colorado, United States as our comparable city due to its similarities in climate. Our model is an example of the potential application of predictive models across various cities to predict flood risk using a city’s physical features.  

## 1.3: Setting Up Libraries
```{r setup, warning = FALSE, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(pscl)
library(plotROC)
library(pROC)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tigris)
library(viridis)
library(ggplot2)
library(dplyr)
```

```{r themes}
mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.75),
  axis.ticks=element_blank())

```

## 1.1: Load Data
```{r load data}
# dependent variable
inundation <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_depe.csv")

# independent variables
elevation <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_elev.csv")

flow_acc <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_fac.csv")

land_use <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_lulc.csv")

slope <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_slope.csv")

water_body <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/calgary_table/zonal_water.csv")

```

# 2: Exploratory Analysis
## 2.1: Fishnet and Joining to show Past Flood Inundation
```{r Fishnet}

calgary_fishnet<-st_read("https://raw.githubusercontent.com/luyiiw/LUEM/main/Calgary_JSON/Calgary_Fishnet.json")
calgary_boundary<-st_read("https://raw.githubusercontent.com/luyiiw/LUEM/main/Calgary_JSON/Calgary_Boundary.json")

calgary_Floods <- calgary_fishnet %>% 
  left_join(inundation, by = "FID")

calgary_Floods$RANGE<-as.factor(calgary_Floods$RANGE)

ggplot() + 
  geom_sf(data=calgary_fishnet,color="grey",size=0.2) +
  geom_sf(data=calgary_Floods ,aes(fill=RANGE),size=0.5) +
   scale_fill_manual(values = c("1"= "#1f78b4", "0" = "#e7f5c6"),
                     labels = c("1" = "Inundation", "0" = "No Inundation"),
                  name = "Inundation Status") +
  labs(title="Past Flood Inundation in Calgary, Canada",
       caption = "Figure 1.1")+
  mapTheme+
    guides(fill = guide_legend(override.aes = list(shape = 22, size = 6)))

```

## 2.2: Data Exploration
```{r}
# Creating tables to combine exploratory data
flood_dep <- inundation %>%
  select(FID, RANGE) %>%
  rename(flood = RANGE) %>%
  mutate(flood = factor(flood))

# 1 = flood, 0 = no flood

#1 Elevation
elev_comb <- elevation %>%
  select(FID, MEAN) %>%
  rename(mean_elev= MEAN)

#2 Slope
slope_comb <- slope %>%
  select(FID, MAX) %>%
  rename(max_slope= MAX)

#3 Water body
water_comb <- water_body %>%
  select(FID, MIN) %>%
  rename(min_dist_water= MIN)

#4 Flow accumulation
fac_comb <-  flow_acc %>%
  select(FID, MEAN) %>%
  rename(mean_fac= MEAN)
#changed from max to mean

#5 Land Use/Permeability
lu_comb <- land_use %>%
  select(FID, MEAN) %>%
  rename(land_use = MEAN) %>%
  mutate(permeability = case_when(
    land_use < 0.25 ~ '0',
    land_use >= 0.25 ~ '1')) %>%
  select(-land_use)

# Removing unnecessary data sets
rm(inundation, elevation, flow_acc, land_use, slope, water_body)

# Combining the tables 
flood_comb <- flood_dep %>%
  inner_join(elev_comb, by = "FID") %>%
  inner_join(slope_comb, by = "FID") %>%
  inner_join(water_comb, by = "FID") %>%
  inner_join(fac_comb, by = "FID")

floodPlotVariables <- 
  flood_comb %>%
  as.data.frame() %>%
  select(flood, mean_elev, max_slope, min_dist_water, mean_fac) %>%
  gather(variable, value, -flood)

```

```{r plot_variables}
# need to check the labels for flood and not flood 
ggplot(floodPlotVariables %>%
         group_by(flood, variable) %>%
         summarize(mean = mean(value))) + 
     geom_bar(aes(as.factor(flood), 
                  mean, 
                  fill=as.factor(flood)),
              stat="identity") + 
     facet_wrap(~variable, scales = "free_y") +
     scale_fill_manual(values = c("#e7f5c6", "#1f78b4"),
                      labels = c("No Flood","Flood"),
                      name = "") +
  labs(title="Continuous Variables Impacting Flood Risk",
       caption = "Figure 2.1")+
    labs(x="Flood", y="Value") +
  plotTheme

```


# 3: Model Building 
## 3.1: Partition Training and Test Sets
```{r}
# Combine all data sets for model building
model_data <- flood_dep %>%
  inner_join(elev_comb, by = "FID") %>%
  inner_join(slope_comb, by = "FID") %>%
  inner_join(water_comb, by = "FID") %>%
  inner_join(fac_comb, by = "FID") %>%
  inner_join(lu_comb, by = "FID") %>%
  mutate(permeability = as.factor(permeability))

#land cover is dummy var and categorical var


# Creating training and test sets
set.seed(3456)
trainIndex <- createDataPartition(model_data$permeability, p = .70,
                                  list = FALSE,
                                  times = 1)

floodTrain <- model_data[ trainIndex,]
floodTest  <- model_data[-trainIndex,]

```


## 3.2: Making the Binomial Model
```{r}
floodModel <- glm(flood ~ ., 
                    family="binomial"(link="logit"), data = floodTrain %>%
                                                            as.data.frame() %>%
                                                            select(-FID))
summary(floodModel)
```

## 3.3: Model Validation
```{r}
classProbs <- predict(floodModel, floodTest, type="response")

hist(classProbs, col = "#1f78b4", main = "Histogram of Flood Probabilities") 


testProbs <- data.frame(obs = as.factor(floodTest$flood),
                        pred = classProbs)

ggplot(testProbs, aes(x = pred, fill=as.factor(obs))) + 
  geom_density() +
  facet_grid(obs ~ .) + 
  xlab("Probability") +
  ylab("Frequency")+
  geom_vline(xintercept = .5) +
  labs(title="Model Validation",
       caption = "Figure 3.2")+
  scale_fill_manual(values = c("#e7f5c6", "#1f78b4"),
                      labels = c("Not Flood","Flood"),
                      name = "")+
  plotTheme
```

## 3.4: Confusion Matrix
```{r confusion_matrix}
testProbs$predClass  = ifelse(testProbs$pred > .5 ,1,0)

caret::confusionMatrix(reference = as.factor(testProbs$obs), 
                       data = as.factor(testProbs$predClass), 
                       positive = "1")
```

Predicted = 0, Observed = 0 —> True Negative (No flood correctly predicted 130)

Predicted = 1, Observed = 1 —> True Positive (Floods correctly predicted 40)

Predicted = 1, Observed = 0 —> False Positive (Incorrectly predicted flood when there was no flood 23)

Predicted = 0, Observed = 1 —> False Negative (Incorrectly predicted no flood when there was a flood 10)

Sensitivity: 0.63
Specificity: 0.92


## 3.5: ROC Curve
```{r roc_curve, message = FALSE, warning = FALSE}
# change obs to numeric
testProbs$obs <- as.numeric(as.character(testProbs$obs))

# plot results
ggplot(testProbs, aes(d = obs, m = pred)) + 
  geom_roc(n.cuts = 50, labels = FALSE) + 
  style_roc(theme = theme_grey) +
  geom_abline(slope = 1, intercept = 0, size = 1.5, color = 'grey') +
  labs(title="ROC Curve",
       caption = "Figure 3.3") +
  plotTheme

#area under the curve
auc(testProbs$obs, testProbs$pred)
```
Area under the curve: 0.85

## 3.6: Cross-Validation
```{r cross_validation}
ctrl <- trainControl(method = "cv", 
                     number = 100, 
                     p = 0.7, 
                     savePredictions = TRUE)

cvFit <- train(as.factor(flood) ~ .,  data = model_data %>% 
                                                as.data.frame(), 
               method="glm", family="binomial",
               trControl = ctrl)

cvFit

ggplot(as.data.frame(cvFit$resample), aes(Accuracy)) + 
  geom_histogram(fill = "#1f78b4", color = "black") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x="Accuracy",
       y="Count") +
  labs(title="Cross-Validation",
       caption = "Figure 3.4")+
  plotTheme

```
Accuracy: 0.86
Kappa: 0.63

<<<<<<< HEAD
## Map Predictions
=======
## 3.7: Goodness-of-Fit
```{r}
allPredictions <- 
  predict(cvFit, model_data, type="prob")[,2]
  
finalflood <- 
  cbind(model_data,allPredictions) %>%
  mutate(allPredictions = round(allPredictions * 100)) 

joined_fishnet_cal <- calgary_fishnet %>% 
  left_join(finalflood, by = "FID")

ggplot() + 
    geom_sf(data=joined_fishnet_cal, aes(fill=factor(ntile(allPredictions,5))), 
            colour=NA) +  # Raster layer
  geom_sf(data = calgary_fishnet, color = "grey", fill = NA) + 
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels=as.character(quantile(joined_fishnet_cal$allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  mapTheme +
  labs(title="Predicted Probability of Flood Inundation in Calgary, Canada")



ggplot() + 
    geom_sf(data=joined_fishnet_cal, aes(fill=factor(ntile(allPredictions,5))), 
            colour=NA) +
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels=as.character(quantile(joined_fishnet_cal$allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  geom_sf(data=joined_fishnet_cal  %>% 
               filter(flood == 1), 
               fill="#de2d26",colour=NA) +
  geom_sf(data = calgary_fishnet, color = "grey", fill = NA) + 
  mapTheme +
  labs(title="Predicted Probabilities and Past Flood Inundation",
       subtitle="Calgary; Past Flood Inundation in red")


##Confusion Matrix Map
joined_fishnet_cal %>%
  mutate(confResult=case_when(allPredictions < 50 & flood==0 ~ "True Negative",
                              allPredictions >= 50 & flood==1 ~ "True Positive",
                              allPredictions < 50 & flood==1 ~ "False Negative",
                              allPredictions >= 50 & flood==0 ~ "False Positive")) %>%
  ggplot()+
  geom_sf(aes(fill = confResult), color = "transparent")+
   geom_sf(data = calgary_fishnet, color = "grey", fill = NA) +
  scale_fill_manual(values = c("Red","Orange","Light Blue","Light Green"),
                    name="Outcomes")+
  labs(title="Confusion Matrix Map") +
  mapTheme
```


# 4: Applying the Model to Denver, Colorado 
## 4.1: Load Data Denver
```{r load_data_den}
# independent variables
elevation_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_elevation_den.csv")

flow_acc_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_fac_den.csv")

land_use_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_permea_den.csv")

slope_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_slope_den.csv")

water_body_den <- read.csv("https://raw.githubusercontent.com/luyiiw/LUEM/main/denver_table/Zonal_waterdist_den.csv")

```

## 4.2: Data Cleaning Denver
```{r data_cleaning_den}
#1 Elevation
elev_comb_den <- elevation_den %>%
  select(FID, MEAN) %>%
  rename(mean_elev= MEAN)

#2 Slope
slope_comb_den <- slope_den %>%
  select(FID, MAX) %>%
  rename(max_slope= MAX)

#3 Water body
water_comb_den <- water_body_den %>%
  select(FID, MIN) %>%
  rename(min_dist_water= MIN)

#4 Flow accumulation
fac_comb_den <-  flow_acc_den %>%
  select(FID, MEAN) %>%
  rename(mean_fac= MEAN)
#changed from max to mean

#5 Land Use/Permeability
lu_comb_den <- land_use_den %>%
  select(FID, MEAN) %>%
  rename(land_use = MEAN) %>%
  mutate(permeability = case_when(
    land_use < 0.25 ~ '0',
    land_use >= 0.25 ~ '1')) %>%
  select(-land_use)

```

## 4.3: Model Building Denver
```{r}
# Combine all data sets for DENVER model building
model_data_den <- elev_comb_den %>%
  inner_join(slope_comb_den, by = "FID") %>%
  inner_join(water_comb_den, by = "FID") %>%
  inner_join(fac_comb_den, by = "FID") %>%
  inner_join(lu_comb_den, by = "FID") %>%
  mutate(permeability = as.factor(permeability))

# Removing unnecessary data sets 
rm(elev_comb_den, slope_comb_den, water_comb_den, fac_comb_den, lu_comb_den)
```
## 4.4: Creating Fishnet Denver
```{r}
denver_fishnet<-st_read("https://raw.githubusercontent.com/luyiiw/LUEM/main/Denver_JSON/denver_fishnet.json")
denver_boundary<-st_read("https://raw.githubusercontent.com/luyiiw/LUEM/main/Denver_JSON/denver_boundary.json")


```
## 4.5: Predicting Flood in Denver
```{r}
den_classProbs <- predict(floodModel, model_data_den, type = "response")

den_floodModel <- cbind(model_data_den, den_classProbs) %>%
  mutate(allPredictions2 = round(den_classProbs * 100)) 

joined_fishnet_den <- denver_fishnet %>% 
  left_join(den_floodModel , by = "FID")


mean(joined_fishnet_den$allPredictions2)

ggplot() + 
    geom_sf(data=joined_fishnet_den, aes(fill=factor(ntile(allPredictions2,5))), 
            colour=NA) +  # Raster layer
  geom_sf(data = denver_fishnet, color = "grey", fill = NA) + 
    scale_fill_manual(values = c("#eff3ff", "#bdd7e7", "#bcaecc", "#5a4565", "#3e204f"),
                      labels=as.character(quantile(joined_fishnet_den$allPredictions2,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  mapTheme +
  labs(title="Predicted Probability of Flood Inundation in Denver, Colorado")
```

